Two Problems for Sophistication

### 2015-07-31 Facticity/Facticity/paper.tex
...

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\Nm}{\mathscr N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\bi}{\bar\imath}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}

@@

\author{Peter Bloem and Steven de Rooij}

\institute{
  System and Network Engineering Group, \\University of Amsterdam, the Netherlands\\
  \email{uva@peterbloem.nl, steven.de.rooij@gmail.com}
}

@@

\section{Introduction}
\enlargethispage{2\baselineskip}
\footnotetext{Peter Bloem is a student author (PhD student).}

@@

The Kolmogorov complexity $C(x)$ of a binary string $x$ is, informally, the length of the shortest computer program to print $x$. This length depends on the choice of programming language, but, by the invariance theorem \cite[Section~2.1]{li1993introduction}, this choice affects the value only by a constant, independent of $x$. For sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure. A definition of sophistication $S(x)$ in the spirit of $C(x)$ should have similar guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except it is also bounded). 
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many proposals for such a measure, all based on a \emph{two-part code}: we encode a \emph{model} in the first part of the code, which is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined.\footnotemark

@@

For any string $x$, there may be many different two-part codes. The total length can never be less than the Kolmogorov complexity, but it can come close. Figure~\ref{fig:diagram} illustrates the principle. The various proposals define sophistication as the size of the model for one of the representations that approaches the Kolmogorov complexity. However, for most definitions, we can prove  they fail one of the conditions above. For certain variants, we cannot \emph{prove} they conflict with our requirements, but we show these methods only assign substantial sophistication to strings that require an enormous amount of processing to construct.

\begin{figure*}[tb]
  \centering
  \begin{minipage}{0.40\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.40\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$ by the two components of their codelength. The Kolmogorov complexity $C^\M(x)$, appearing as a black diagonal, provides a lower bound on the total codelength. We consider only representations that are close to this optimum, with the threshold represented by a dashed line. The size of the smallest model below the threshold is the sophistication of the data. (right) The same image, after a constant perturbation in the model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

A valid definition of $S$ must contend with two important issues. First, the details of the way the model is encoded are important. There are two technically distinct approaches; in one of these one has to deal with the so-called ``nickname problem'' that strangely remains unresolved in several publications. These definitions yield a sophistication that is highly dependent on the chosen programming language, unless special care is taken, as discussed in Section~\ref{section:indices}.

@@

Underfitting occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is also a problem for sophistication because the models under consideration are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} are \emph{simple}, since they can be described with a relatively small number of bits, yet are able to represent the data using a code within a constant from the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models, for instance to total functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, In the discussion in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as poor representations, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable representations. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that the data allows for multiple equivalent interpretations of which information is structured, and which is random, and that there is no such thing as sophistication.

\enlargethispage{3\baselineskip} 
\section{Notation}
The following notation allows us to generalize across all definitions and variants, save the occasional exception which we will highlight individually.

Let $\B = \{0,1\}^*$. We deal with partial computable functions $f: \B \times \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\text{dom}_z(f)~=~\{y~:~f(y,~z)~\neq \infty\}$, is a prefix free set for all $z$, i.e. no string in $\text{dom}_z(f)$ is a prefix of another. A function $f$ is \emph{total} if $\forall_z \text{dom}_z(f) = \B$. In most cases, we do not use the second argument, and we adopt the convention that $f(x) = f(x, \epsilon)$.

A \emph{numbering} is an enumeration of of the partial computable functions, denoted as $\psi_1, \psi_2, \ldots$ or simply $\psi$. We fix one canonical effective numbering $\phi$. We call a numbering $\psi$ \emph{acceptable} if there exist total, computable functions $a, b: \N \to \N$ with $\forall: i$, $\phi_i = \psi_{b(i)}$ and  $\psi_i = \phi_{a(i)}$.

A \emph{model class} is a set of model indices. We distinguish four classes:
\begin{itemize} 
  \item The indices of the partial computable functions $\C=\N$.
  \item The total functions $\T=\{i:\tn{$\phi_i$ is total}\}$. Note that $\T$ is not computably enumerable.
  \item $\K$ is an enumerable set such that $\{\phi_i:i\in\K\}$ is the set of all partial computable prefix functions.
  \item The finite sets: $\F$ is an enumerable set such that $\{\phi_i:i\in\F\}$ contains a uniform code for every finite set.\footnote{A uniform code for a set $S$ is a surjective prefix function $f:\{0,1\}^{\lceil\log|S|\rceil}\to S$.}
\end{itemize}
A model class is thus defined in reference to a specific numbering, indicated by the context.

We sometimes need to map binary strings to a prefix-free set. We denote by $\br{x}$ the prefix-encoded representation for $x$. We require that the mapping satisfies $|\br{x}| = |x|+O(\log(|x|)$ (see eg. \cite[Section~1.4]{li1993introduction}). To simplify notation, we will sometimes conflate natural numbers and binary strings, implicitly using the ordering $(0, \epsilon)$, $(1, 0)$, $(2, 1)$, $(3, 00)$, $(4, 01)$, \ldots

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: we define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x\mid z)=\min\{\bar\imath y:\psi_i(y; z)=x,i\in\M\}$, with $C^{\M,\psi}(x) = C^{\M,\psi}(x\mid \epsilon)$. We may drop the numbering parameter when the distinction isn't relevant. Under this notation $C^\C(x)$ corresponds to the plain Kolmogorov complexity $C(x)$ and $C^\K(x)$ corresponds to the prefix-free version $K(x)$. We can also represent the smallest two-part description of $x$ using a given model $\phi_i$ as $C^{\{i\}}(x)$. The complexity of a finite set is defined an the complexity of an explicit enumeration.

We use a choice of numbering for the purpose that is normally served by the universal Turing machine. The choice is a matter of taste: a UTM determines a numbering, and vice versa. The reason that we prefer to work with numberings is that it highlights an important issue: while Kolmogorov complexity is invariant to the choice of numbering this property does not immediately carry over to sophistication: for some treatments, the result is highly dependent on the chosen numbering (as shown in the next section).

@@

The simplest approach to sophistication would be to `open up' the Kolmogorov complexity and to see which program on the universal Turing machine achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. Since any UTM uses a two-part coding, its programs consist of a model and an input:

\begin{definition}[Index sophistication]
Let $\psi$ be an acceptable numbering. Let $\M$ be the model class from which candidates are chosen, and let $\Nm$ be the model class that determines the minimum achievable complexity. Let $c$ be a fixed constant. The \emph{index sophistication} is:
\[
\s_{\text{index}}^{\M,\Nm,\psi,c}(x) = \min\left\{ |i| : i \in \M, C^{\{i\}, \psi}(x) \leq C^{\Nm, \psi}(x)+c   \right\} \p
\]
When $\M = \Nm$, we will abbreviate with $\s^{\Nm,\psi,c}$. \label{definition:index}
\end{definition}
Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains idiosyncracies like the use of monotonic models, and an extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication,antunes2013sophistication} Koppel's principle is limited to finite strings, with the total functions as a model class. The definition used is similar to $s_\text{index}^{\T, \C,\psi,c}$, with one exception: the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This difference is not relevant to the point we make in this section. The restriction to total classes is a common approach, perhaps used because it helps to avoid underfitting, which we will discuss in the next section.

\begin{lemma}
Let $\s^\psi$ denote any index sophistication with respect to numbering $\psi$.
There are acceptable numberings $\psi$ and $\xi$ such that for all $x$:\\
\-\hspace{2.8em} $|\s^\psi(x) - \s^\xi(x)|\geq \tfrac{1}{2}\min\{\s^\psi(x),\s^\xi(x)\}$.
\end{lemma}
\begin{proof}
Let $s_i$ be the string consisting of $2^{i}$ zeroes followed by a one. Define $\psi, \xi$ such that $\psi_j(x) = \phi_i(x)$ for $j = s_{2i}$ and $\xi_j(x) = \phi_i(x)$ for $j = s_{2i+1}$, with all other functions returning $\infty$ for all inputs. Choose any $x$ and assume w.l.o.g. that $\s^\psi(x)\le s^\xi(x)$. By construction, we have $2\s^\psi(x)  \leq \s^\xi(x)$.
\end{proof}
Thus, the length of the index is a very poor indicator of model complexity. For a robust measure, define the complexity of functions as in \cite{grunwald2004shannon,vitanyi2004meaningful} by $C^{\M,\psi}(f) = \min\{C^{\M,\psi}(i):\phi_i=f\}$.
Lemma~\ref{lemma:invariance} in the appendix shows that $C^\C(f)$ and $C^\K(f)$ are invariant: they are objective properties of a function.

Note that such perversely inefficient numberings are not an issue for Kolmogorov complexity. If our UTM uses an inefficient numbering, we can switch to a more efficient one at only a constant cost, keeping the complexity the same, up to a constant. For sophistication, however the choice of numbering is crucial.

There are two ways to use the model complexity $C^\M(\phi_i)$ for more robust attempts to define sophistication. Confusingly, both are used in the literature. First, we can measure the complexity of the \emph{model} using $C^{\M,\psi}(\phi_i)$, which is then the size of the first part of a two-part code describing the data. For instance, this approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information} and in most of the present paper.

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings. This approach is taken by Adriaans in \cite{adriaans2012facticity}, which defines \emph{facticity} as $\s_\text{index}^{\C,\psi,0}$, but only allows numberings that are \emph{faithful}: $\forall i \exists j : \psi_i=\psi_j, |j|\le C^{\C}(\psi_j)+c$, for some constant $c$.

We prove in the appendix that---contrary to Adriaans' suggestion---there actually do exist faithful, acceptable numberings (Lemma~\ref{lemma:faithful-numberings}). However even choosing a faithful index is not quite enough. The Kolmogorov complexity uses representations of the form $\bar\imath y$, where the bar denotes some straightforward prefix encoding to delimit the model description $i$ from its input $y$. If we define a second prefix encoding $\tilde{\imath}$, with $|\bar\imath|-|\tilde\imath|$ unbounded, we can define a second representation $\bar u \tilde \imath y$, at a constant overhead $|\br{u}|$, and gain more than $|\br{u}|$ for sufficiently complex strings, resulting in a bounded sophistication.

We continue with a sophistication that avoids these issues. Let $C'^{\M, \psi} = C^\K(\psi_i) + min\{|y| : \psi_i(y) = x\}$. We define sophistication as:

\begin{definition}[Sophistication]
\[
\s^{\M,\Nm,\psi,c}(x)=\min\left\{C^\K(\phi_i):i\in\M,C'^{\{i\},\psi}(x)\le C^{\Nm,\psi}(x)+c\right\}.
\]
\end{definition}

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the variant $\s^{\K,\psi,c}$. This variant itself is not used anywhere in the literature, but it helps to illustrate the issues we wish to discuss.

$\K$ has representations with all information except a constant in the input and representations with all information in the model. The downside to this balance is that it becomes easy to show a lack of invariance: by tweaking the numbering so that models in a specific subset $\M'\subset\M$ become cheaper to represent by an arbitrary amount relative to others, so that a model in $\M'$ always determines the sophistication.

This means that for model classes with a universal model, we can choose our numbering to make $S^\K$ either bounded, or equal to the complexity:

\begin{theorem}[Underfitting]
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M, x : C'^{\{u\}}(x) \leq C'^{\{i\}}(x) + c$. Then, for some numberings $\s^{\M,\psi,c}$ is bounded. \label{theorem:underfitting}
\end{theorem}

This result is well known and many treatments avoid it by restricting the model class. Less well known is perhaps that the same holds in the other direction: for some numberings the singletons always determine the sophistication.

\begin{theorem}[Overfitting]
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\M(x)-S^{\M,\psi,c}(x)\leq c$.\label{theorem:overfitting}
\end{theorem}
The proofs of both theorems rely on a simple principle: for certain $\M' \subset \M$, there exist numberings which have the effect of penalizing $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this to effectively `push' these models outside of the range of candidates, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication.

For these theorems, this means setting $\M' = \{u\}$, with $\phi_u$ a universal model, or letting $\M'$ be the singletons. The requirements for $\M'$ are strict, and slightly complex. The following lemma gives a set of sufficient conditions.

\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\M$ be any model class, let $\X$ be any set of binary strings and let $D:\B\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\phi$. Let $\M'= \tn{range}(D)$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1) $\forall_{m\in\M'}:\min\{|p|:\phi_{D(p)}=\phi_m\}\le C^{\K,\phi}(\phi_m)+c$\\
\-\hspace{1cm}(2) $\forall_{x\in\X}:C'^{\M',\phi}(x)-C'^{\M,\phi}(x)\le c$.\\
Then there is an enumeration $\psi$ of the partial recursive functions such that $S^{\M,\psi,c}(x) = |\br{0}|+S^{\M',\phi,c}(x)$ for all $x\in\X$.
\end{restatable}

>>>>

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\Nm}{\mathscr N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\bi}{\bar\imath}
\newcommand{\dom}[1]{\mathop{\tn{dom}(#1)}}
\newcommand{\range}[1]{\mathop{\tn{range}(#1)}}
\newcommand{\Cc}{C_\tn{comp}}

@@

\author{Peter Bloem, Steven de Rooij and Pieter Adriaans}

\institute{
  System and Network Engineering Group, \\University of Amsterdam, the Netherlands\\
  \email{uva@peterbloem.nl, steven.de.rooij@gmail.com, p.w.adriaans@uva.nl}
}

@@

\section{Introduction}\enlargethispage{3\baselineskip}

@@

The Kolmogorov complexity $C(x)$ of a binary string $x$ is, informally, the length of the shortest computer program to print $x$. This length depends on the choice of programming language, but, by the invariance theorem \cite[Section~2.1]{li1993introduction}, only by a constant, independent of $x$. For sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure. A definition of sophistication $S(x)$ in the spirit of $C(x)$ should have similar guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except it is also bounded). 
\item Similarly, there should be no constant $c$ such that $\left | C(x)-S(x)\right |\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many proposals for such a measure, all based on a \emph{two-part code}: we encode a \emph{model} in the first part of the code, which is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined.\footnotemark

@@

For any string $x$, there may be many different two-part codes. The total length can never be less than the Kolmogorov complexity, but it can come close. Figure~\ref{fig:diagram} illustrates the principle. The key to sophistication is to take the representations that come close to the Kolmogorov complexity, the \emph{candidates}, and define the sophistication as the size of the smallest model in this set. However, for most definitions, we can prove that they fail one of the conditions above. For others, we cannot \emph{prove} they conflict with our requirements, but we show these methods only assign substantial sophistication to strings that require an enormous amount of processing to construct.

\begin{figure*}[tb]
  \vspace{-2\baselineskip}
  \centering
  \begin{minipage}{0.40\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.40\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$ by the two components of their code. The Kolmogorov complexity $C(x)$, appearing as a black diagonal, provides a lower bound on the total codelength. We consider only representations that are close to this optimum---called \emph{candidates}---with the threshold represented by a dashed line. The size of the smallest model below the threshold is the sophistication of the data. (right) The same image, after a constant perturbation in the model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}\enlargethispage{3\baselineskip}

A valid definition of $S(x)$ must contend with two important issues. First, the details of the way the model is encoded are important. There are two technically distinct approaches; in one of these one has to deal with the so-called ``nickname problem'' that strangely remains unresolved in several publications. These definitions yield a sophistication that is highly dependent on the chosen programming language, unless special care is taken, as discussed in Section~\ref{section:indices}.

@@

Underfitting occurs when the selected model is simple, but fails to capture all structure in the data. This is also a problem for sophistication because the models under consideration are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} are \emph{simple}, since they can be described with a relatively small number of bits, yet are able to represent any data using a code within a constant from the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models, for instance to total functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and identifies some models as poor representations, it is probably not possible to objectively separate structure from noise and identify a \emph{single} model as ``best'': many models of different complexities may be reasonable representations. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that the data allows for multiple, equivalent interpretations of which information is structured, and which is random, and that there is no such thing as sophistication.

\section{Notation}\enlargethispage{3\baselineskip}
The following notation allows us to generalize across all definitions and variants, save the occasional exception which we will highlight individually.

Let $\B = \{0,1\}^*$. We deal with partial computable functions $f: \B \times \B \to \B$, which we
also call \emph{models}. $f$ is called \emph{prefix} if $\text{dom}_z(f)~=~\{y~:~f(y,~z)~\neq \infty\}$, is a prefix free set for all $z$, i.e. no string in $\text{dom}_z(f)$ is a prefix of another. A function $f$ is \emph{total} if $\forall_z \text{dom}_z(f) = \B$. In most cases, we do not use the second argument, and let $f(x) = f(x, \epsilon)$.

A \emph{numbering} is an enumeration of the partial computable functions, denoted by $\psi_1, \psi_2, \ldots$ or simply $\psi$. We fix one canonical numbering $\phi$, chosen to be \emph{effective}: ie. given $i$ and $y$, we can effectively compute $\phi_i(y)$. We call a numbering $\psi$ \emph{acceptable} if there exist total, computable functions $a, b: \N \to \N$ with $\forall: i$, $\phi_i = \psi_{b(i)}$ and  $\psi_i = \phi_{a(i)}$.

A \emph{model class} is a set of indices in a numbering $\psi$. We define four classes:
\begin{itemize} 
  \item The indices of the partial computable functions $\C=\N$.
  \item The total functions $\T=\{i:\tn{$\psi_i$ is total}\}$. Note that $\T$ is not computably enumerable.
  \item $\K$ is an enumerable set such that $\{\psi_i:i\in\K\}$ is the set of all partial computable prefix functions.
  \item The finite sets: $\F$ is an enumerable set such that $\{\psi_i:i\in\F\}$ is the set of uniform codes for all finite sets.\footnote{A uniform code for a set $F$ is a surjective prefix function $f:\{0,1\}^{\lceil\log|F|\rceil}\to F$.}
\end{itemize}

Let $\br{x}$ denote the prefix-encoded representation for $x$. We require that the mapping satisfies $|\br{x}| = |x|+O(\log|x|)$ (see eg. \cite[Section~1.4]{li1993introduction}). To simplify notation, we will sometimes conflate natural numbers and binary strings, implicitly using the ordering $(0, \epsilon)$, $(1, 0)$, $(2, 1)$, $(3, 00)$, $(4, 01)$, \ldots

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: let $\M$ be a model class and $\psi$ an acceptable numbering, then let $C^{\M,\psi}(x\mid z)=\min\{|\bar\imath y|:\psi_i(y, z)=x,i\in\M\}$, with $C^{\M,\psi}(x) = C^{\M,\psi}(x\mid \epsilon)$. We omit the numbering when the distinction is not relevant. $C^\C(x)$ corresponds to the plain Kolmogorov complexity $C(x)$ and $C^\K(x)$ corresponds to the prefix-free version $K(x)$. Note that the notation $C^{\{i\}, \psi}(x)$ represents the smallest two-part description of $x$ using model $\psi_i$.

We use the principle of a numbering for the purpose normally served by the universal Turing machine. We prefer to work with numberings as it highlights an important issue: while Kolmogorov complexity is invariant to the choice of numbering this property does not immediately carry over to sophistication: for some treatments, the result is highly dependent on the chosen numbering, as we will see in the next section.

@@

The simplest approach to sophistication would be to `open up' the Kolmogorov complexity and to see which program achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. This witness is a two-part coding; it consists of a model and an input.

\begin{definition}[Index sophistication]
Let $\psi$ be an acceptable numbering. Let $\M$ be the model class from which candidates are chosen, and let $\Nm$ be the model class that determines the minimum achievable complexity. Let $c$ be a fixed constant. The \emph{index sophistication} is:
\[
\s_{\tn{index}}^{\M,\Nm,\psi,c}(x) = \min\left\{ |i| : C^{\{i\}, \psi}(x) \leq C^{\Nm, \psi}(x)+c,\;i \in \M\right\} \p
\]
When $\M = \Nm$, we will use $\s_\tn{index}^{\M,\psi,c}$. If the set over which the minimum is taken is empty, the sophistication is undefined.\label{definition:index}
\end{definition}
Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains idiosyncracies like the use of monotonic models, and an extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication,antunes2013sophistication} Koppel's principle is limited to finite strings, with $\T$ as a model class. The definition is similar to $S_\tn{index}^{\T, \C,\psi,c}$, except the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This difference is not relevant to the current discussion. The restriction to $\T$ is a common approach, which avoids underfitting, as discussed in the next section.

\begin{lemma}\label{lemma:perverse}
Let $\s_\tn{index}^\psi$ denote any index sophistication with respect to numbering $\psi$ (with any choice for $\M$, $\Nm$ and $c$). There are acceptable numberings $\psi$ and $\xi$ such that for all $x$: $|\s_\tn{index}^\psi(x) - \s_\tn{index}^\xi(x)|\geq \tfrac{1}{2}\min\{\s_\tn{index}^\psi(x),\s_\tn{index}^\xi(x)\}$.
\end{lemma}
\begin{proof}
Let $z_i \in \B$ consist of $2^{i}-1$ zeroes followed by a one. Define $\psi, \xi$ such that $\psi_j(x) = \phi_i(x)$ for $j = z_{2i}$ and $\xi_j(x) = \phi_i(x)$ for $j = z_{2i+1}$, with all other functions returning $\infty$ for all inputs. Choose any $x$ and assume w.l.o.g. that $\s_\text{index}^\psi(x)\le \s_\text{index}^\xi(x)$. By construction, we have $2\s_\text{index}^\psi(x)  \leq \s_\text{index}^\xi(x)$.
\end{proof}\enlargethispage{3\baselineskip}
Thus, the length of the index is a very poor indicator of model complexity. For a robust measure, we define the complexity of a function $f$ as in \cite{grunwald2004shannon,vitanyi2004meaningful} by 
\begin{align}
C^{\M,\psi}(f) = \min\{C^{\M,\psi}(i):\psi_i=f\} \p \label{equation:function-complexity}
\end{align}
Lemma~\ref{lemma:invariance} in the appendix shows that $C^\C(f)$ and $C^\K(f)$ are invariant. Note that the perversely inefficient numberings of Lemma~\ref{lemma:perverse} are no issue for Kolmogorov complexity: we can use a UTM with a more efficient numbering as a model at only a constant penalty. For sophistication, however, the numbering is crucial.

There are two ways to use $C^\M(f)$ for more robust attempts to define sophistication. Confusingly, both are used in the literature. First, we can measure the complexity of the \emph{model} $\phi_i$ as $C^\K(\phi_i)$, which is then the size of the first part of a two-part code describing the data. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information}.

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings to those that can represent a given function \emph{efficiently}. This approach is taken by Adriaans in \cite{adriaans2012facticity}, who defines \emph{facticity} as $\s_\text{index}^{\C,\psi,0}$, but only allows \emph{faithful} numberings. Formally, a faithful numbering has the property that $\forall i \exists j : \psi_i=\psi_j, |j|\le C^{\C}(\psi_j)+c$, for some constant $c$. Essentially, this means that a faithful numbering can represent a function $f$ with an index the same length as the Kolmogorov complexity $C^\C(f)$.

We prove in the appendix that---contrary to Adriaans' suggestion---there do exist faithful, acceptable numberings (Lemma~\ref{lemma:faithful-numberings}). However, even choosing a faithful numbering is not enough. The Kolmogorov complexity uses representations of the form $\bar\imath y$, with $\psi_i(y) = x$,  where the bar denotes some straightforward prefix encoding to delimit the model description $i$ from its input $y$. If we define a second prefix encoding $\tilde{\imath}$, with $|\bar\imath|-|\tilde\imath|$ unbounded, we can define a second representation $\bar u \tilde \imath y$, with $\psi_u(\tilde{\imath} y) = \psi_i(y)$, at a constant overhead $|\br{u}|$, and gain more than $|\br{u}|$ for sufficiently complex strings, resulting again in a bounded sophistication.

We continue with a sophistication that avoids the issues of inefficient indices and of inefficient prefix encodings. We change the definition of index sophistication so that its two-part representations use $C^\K(\phi_i)$ bits for the representation of the model. We first introduce the following notation for the $\M$-Kolmogorov complexity using such compact two-part representations:
\[
\Cc^{\M, \psi}(x) = \min\{ C^{\K, \psi}(\psi_i) + |y| :  \psi_i(y) = x, i \in \M\}\p
\]
For model classes $\K$ and $\C$ this is equivalent to the existing definition and invariant to the numbering. Note that again, we use $\Cc^{\{i\}, \psi}(x)$ to represent the smallest two-part code using model $\psi_i$.

\begin{definition}[Sophistication]
\[
S^{\M,\Nm,\psi,c}(x)=\min\left\{C^\K(\phi_i): \Cc^{\{i\}, \psi} \le \Cc^{\Nm,\psi}(x)+c, i\in\M\right\}.
\]
\end{definition}

@@

In the last section, we began to see the delicate balance between the two code components. We will study this balance, starting with the variant $S^{\K,\psi,c}$, which is not used in the literature, but helps to illustrate the issues we wish to discuss.

$\K$ has optimal representations with all but a constant part of the information in the input and it has optimal representations with all information in the model. The downside to this balance is that it becomes easy to show a lack of invariance. We can tweak the numbering so that models in a specific subset $\M'\subset\K$ become cheaper to represent by an arbitrary amount relative to others: we can ensure that a model in $\M'$ always determines the sophistication. For instance, if we let $\M'$ contain only a universal model we get a bounded sophistication.

\begin{theorem}[Underfitting]
Let $\M, \Nm$ be model classes with $\M \subseteq \Nm$ and let $\M$ contain a universal model $\phi_u$, with the property that $\exists c \forall i \in \Nm, x \in \B : \Cc^{\{u\},\phi}(x) \leq \Cc^{\Nm, \phi}(x)+ c$. Then, for some numbering $\psi$, $S^{\M,\Nm, \psi,c}$ is bounded. \label{theorem:underfitting}
\end{theorem}
This problem is well known and many treatments avoid it by restricting the model class. Less well known, perhaps, is that the same holds in the other direction:  if $\M'$ is the set of \emph{singleton} models---those models that output a single $x$ for an empty input---we get a sophistication equal to the Kolmogorov complexity.

\begin{theorem}[Overfitting]
Let $\X \subseteq \B$. Let $\M \subseteq \Nm \subseteq \K$ be model classes where for every $x\in\X$ there is a singleton model $i\in\M$ with $\phi_i(\epsilon)=x$. Then there is a numbering $\psi$, and a constant $c$, such that for all $x\in\X$ we have $C^\K(x)-S^{\M,\Nm,\psi,c}(x) \leq c$.\label{theorem:overfitting}
\end{theorem}
The proofs of both theorems rely on a simple principle: there exist numberings which have the effect of penalizing $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this to effectively `push' these models outside of the range of candidates, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication. The requirements for $\M'$ are somewhat complex. The following lemma gives a set of sufficient conditions.

\begin{lemma}
\label{lemma:thecoolone}
Let $\M$ and $\Nm$ be any model class, let $\X$ be any set of binary strings and let $D:\B\to\N$ be a partial computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\phi$. Let $\M'= \tn{range}(D)$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(a) $\forall_{m\in\M'}:\min\{|p|:\phi_{D(p)}=\phi_m\}\le C^{\K,\phi}(\phi_m)+c$\\
\-\hspace{1cm}(b) $\forall_{x\in\X}: \Cc^{\M', \phi}(x) - \Cc^{\Nm, \phi}(x) \le c$.\\
Then, there is a $\psi$ such that if $S^{\M,\Nm,\psi,k}(x)$ is defined, then $S^{\M,\Nm,\psi,k}(x) = S^{\M',\Nm,\psi,k}(x)$ up to a constant. 
\end{lemma}
\begin{proof}\enlargethispage{2\baselineskip}
Pick any $x \in \X$. Let $f$ and $g$ be $\phi$-indices such that $f \in \M'$ and $g \notin \M'$ nor is $\phi_g$ equivalent to any function indexed by $\M'$. Furthermore let $\Cc^{\{f\}, \phi}(x)$ and $\Cc^{\{g\}, \phi}(x)$ both be within a constant $q$ of $\Cc^{\Nm,\phi}(x)$. Assumption (b) ensures that $\M'$ always provides such an $f$.
